Paper Proposal for POLSCI 688

### 2019-11-28 paper/paper-draft.tex
thanksgiving writing

@@

\begin{document}
\begin{titlepage}
\maketitle
\end{titlepage}

@@

\subsection*{Building and testing theories of preemption}
The prior section's narrative history of preemption suggests that preemption is driven by three actors (localities, states, and interest groups) and their relations to one another. In this section I turn to theorizing about when and why we should expect certain relations of cities, states, and interest groups to result in preemption laws. I structure my discussion around three conditions; a state is most likely to preempt a specific municipal regulation when the following are met:

\singlespacing
\begin{enumerate}
	\item \textit{At least one city in the state is able to pass, or has already passed, the regulation.}
	\item \textit{There is a wide divergence in policy preferences between at least one city in the state and the state.}
	\item \textit{Relevant interest groups are organized at the state level.}
\end{enumerate}
\setstretch{2}

The first condition concerns the threat posed by local cities given their institutional capacity to pass regulations. Some states' legal regimes governing city authority empower local policymaking to a greater extent than others, resulting in a greater likelihood of city action. If cities cannot pass regulations in the first place, states should not be expected to take the time to preempt specific municipal ordinances. The second condition concerns the ideological gap between cities and their state. If cities and states possess different preferences over the given regulatory measure, we should expect states to be more willing to preempt local authority. In states where city and state preferences are closer to one another, states will be more willing to let localities engage in policymaking. Finally, the third condition concerns the activity of relevant interest groups. The narrative above suggests that tobacco lobbying firms, the NRA, and ALEC have played a key role in mobilizing states to clamp down on local policymaking. Where these groups are more powerful, as I detail further below, we should see more preemption activity.

@@

I hope to show that these conditions should apply to all preemption activity, regardless of the policy area. Conservative states that contain cities able to pass minimum wage ordinances with mobilized business groups should be more likely to preempt local minimum wage increases. Likewise, conservative states with active tobacco lobbying firms and cities capable of regulating tobacco consumption should be more likely to preempt local tobacco regulation. While the broad applicability of the hypotheses is an advantage, the reality of the data on state and local preemption and policymaking means we need to temper any hope of testing hypotheses in multiple policy domains. Collecting data on cities' policymaking, which is necessary to test whether the realization of the threat of local action spurs states to act preemptively, is no small task. I only know of one resource, the U.S. Tobacco Control Laws Database (TCLD) maintained by the American Nonsmokers' Rights Foundation,\footnote{\url{https://no-smoke.org/materials-services/lists-maps/}} that tracks both local ordinances and state preemption laws. While collecting data on cities' attempts to pass other regulations is important (especially in the the policy areas that have seen the sharpest rise in preemption activity in recent years), my aims are more modest. I hope to use the TCLD's historic data on local tobacco regulation and state preemption over the past 40 years to test whether my three conditions hold. In the following pages, I provide further justification of the three conditions and offer concrete means of testing them in the context of tobacco preemption.

The first condition concerns cities' capacities to pass regulations. As described above, cities have often lacked this power. Early formulations of Home Rule, paired with narrow interpretations of cities' sphere of policymaking authority, hindered cities' ability to legislate as they saw fit. Since mid-century, many states have adopted legislative Home Rule regimes that permit a greater degree of state policymaking while allowing state courts and legislatures to intervene and deem local ordinances in conflict with state law. Today, while legislative Home Rule predominates, cities are governed by a patchwork of state legal regimes, with different implications for the likelihood of state preemption. In states where Dillon's Rule is still frequently employed by judges, cities possess limited policymaking authority. They can only legislate in policy domains that have been explicitly permitted by the state. While states can theoretically grant wide authority to cities and then preempt ordinances by curtailing city's delegated power, states that employ Dillon's Rule usually delegate narrow powers to cities in the first place. As a result, cities have little capacity to legislate and if they act outside the scope of their delegated authority, it usually falls to the state courts to determine if the ordinance is legal \parencite{dillerIntrastatePreemption2007}.

In other states, cities' authority is structured according to the earlier formulation of Home Rule. Cities enjoy authority over matters of local concern, but the bounds of "local concern" is up to state judges' interpretation. Consequently, if a city ordinance is preempted, it is usually the result of a court deciding that the city acted outside of its local area of concern \parencite{dillerIntrastatePreemption2007}. In contrast, legislative Home Rule empowers cities to pass any ordinance that does not conflict with state law. While cities enjoy broad authority in such a regime, state legislatures, as discussed above, also possess the authority to intervene and curtail the bounds of local autonomy. Therefore, it is in states that govern cities with legislative Home Rule that we should expect to see the greatest level of preemption activity.

Unfortunately, given the complexities of state-local relations, there is no clear method for classifying states' legal regimes. Even the distinction between Home Rule and Dillon's Rule has been challenged \parencite{richardsonDillonRuleMars2011}. In a review of different classification schemes, \textcite{dillerIntrastatePreemption2007} notes that some scholars put the number of states with Home Rule at 48 while others count 45. Differentiating between the type of Home Rule is even trickier. One study counts 26 states with legislative Home Rule while others argue that the vast majority of Home Rule states employ legislative Home Rule. Classification disagreements are not always just at the margin either. In their analysis of the causes of state preemption, \textcite{fowlerStatePreemptionLocal2019}, citing \textcite{kraneHomeRuleAmerica2000}, only treat 10 states as using Home Rule.

Classifying the legal regimes governing cities in the states is a challenging but necessary step for measuring how the threat of local action affects states' decisions to preempt local ordinances. While states' legal regimes provide one perspective on the threat of local action, another more concrete measure of the potential for local action is whether a city in a state has already passed a tobacco regulation. Unlike the state's legal rules governing cities, which measures the potential for action, whether a city has already regulated the sale or consumption of tobacco captures the realization of the threat. Controlling for cities' policymaking is notably absent from recent quantitative analyses of preemption \parencite{fowlerStatePreemptionLocal2019,goodmanStateLegislativeIdeology2019,flavinExplainingStatePreemption2019}, likely due to the difficulty of collecting data on cities' policies. Luckily, the TCLD systematically tracks local tobacco regulations alongside states' preemption laws. It includes the tobacco regulations of 98\% of the cities in the U.S. with more than 75,000 inhabitants, allowing for me to test my theory across all fifty states.

The second condition increasing the likelihood of state preemption is the ideological gap between a state and its cities. Usually, struggles between cities and states pit conservative state governments against liberal cities [CITE THINGS HERE], but this is not always the case.\footnote{For instance, in 2013, Oregon preempted cities from regulating genetically-modified crops and, in 2017, New York overturned New York City's tax on plastic bags \parencite{mckinleyCuomoBlocksNew2017,scharffHyperPreemptionReordering2018}.} Ideology and partisanship play a key role in recent studies of preemption, with Republican control of state government and state conservatism often correlating with state preemption [CITE]. However, the lack of local data has prevented scholars from examining whether Republican control of government and conservatism inherently lead to preemption or whether the connection depends on cities' ideology.

Measuring city ideology is difficult (especially over decades), but Figure \ref{fig:ideology} below offers an attempt at measuring the ideological divide between states and cities. The dashed lines in the figure show the increasing relative support Democratic presidential candidates enjoy in the three largest counties in states that support the Democratic or Republican candidate. Over time, cities have become increasingly more liberal on average than their states. This is especially true for cities in states that support the Republican presidential candidate (red dashed line).

\begin{figure}[ht]
\caption{Average relative support of Democratic presidential candidate by geography and state partisanship, 1980-2016}
\label{fig:ideology}
\centering
\includegraphics[width=.75\textwidth]{plots/county_support}
\newline\scriptsize{This plot shows the average support for the Democratic presidential candidate in and outside the three largest counties in a state relative to the candidate's support in the rest of the state. For each state and year, the democratic candidate's vote share in and outside each state's three largest counties is divided by his or her vote share in the entire state. Annual average measures of relative support are then calculated, grouping by whether the state supported the Democratic candidate and whether the counties are among the three largest in size. Given time constraints, the total number of votes in a county is used as a proxy for the total population of the county. Source: \textcite{cqpressCQVotingElections2019}.}
\end{figure}

The growing ideological divergence between urban and rural areas has been well-documented elsewhere \parencite[e.g.][]{chenUnintentionalGerrymanderingPolitical2013} and cited as a cause of increasing levels of state preemption \parencite{dupuisCityRightsEra2018,hicksHomeRuleBe2018}. As Democratic voters cluster in urban areas, they increase the liberalism of cities relative to states and reduce the number of state-level Republicans representing urban areas. As a result, cities may be more willing to enact more liberal policies that increasingly diverge from the interests of state elected officials, increasing the likelihood that the state will preempt the progressive local regulations.

The clustering of Democratic voters in urban areas

\textcite{chenUnintentionalGerrymanderingPolitical2013} note how Democratic voters are clustering in urban areas that are

IS TOBACCO A PARTISNA ISSUE?

\begin{figure}[ht]
\caption{Number of state governments under unified partisan control, 1980-2017}
\centering
\includegraphics[width=.75\textwidth]{plots/party_control}
\newline\scriptsize{Note: Excludes Nebraska. Sources: \textcite{jordanCorrelatesStatePolicy2017,nationalconferenceofstatelegislaturesStatePartisanComposition2019}}
\end{figure}

\newpage
\singlespacing
\printbibliography

>>>>

\usepackage[colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black]{hyperref}

@@

\begin{document}
\begin{titlepage}
\maketitle
\end{titlepage}
\hypersetup{pageanchor=false}

@@

\subsection*{Building and testing theories of preemption}
The prior section suggests that preemption has been driven by three actors (localities, states, and interest groups) and their relations to one another. In this section, I drawn upon the history of preemption and the existing literature to develop hypotheses about about when and why we should expect certain relations of cities, states, and interest groups to result in preemption laws. I argue that a state is most likely to preempt a specific municipal regulation when the following three conditions are met:

\singlespacing
\begin{enumerate}
	\item \textit{At least one city in the state is able to pass, or has already passed, the regulation.}
	\item \textit{There is a wide divergence between the policy preferences of the state government and at least one city in the state.}
	\item \textit{Relevant interest groups are organized at the state level.}
\end{enumerate}
\setstretch{2}

The first condition concerns the threat of local policymaking. The legal regimes governing cities' authority empower local policymaking to a greater extent in some states than in others, resulting in a greater likelihood of municipal action. If cities cannot pass regulations in the first place, states should not be expected to take the time to preempt specific municipal ordinances. The second condition concerns the ideological gap between cities and the state. If cities and states possess different preferences over the given regulatory measure, we should expect states to be more willing to preempt local authority. In states where city and state preferences are closer to one another, states will be more willing to let localities engage in policymaking. Finally, the third condition concerns the activity of relevant interest groups. The narrative above suggests that tobacco lobbying firms, the NRA, and ALEC have played a key role in mobilizing states to clamp down on local policymaking. Where these groups are more powerful, we should see more preemption activity.

@@

I hope to show that these conditions should apply to all preemption activity, regardless of the policy area. While the broad applicability of the hypotheses is an advantage, the reality of the data on state preemption and local policymaking means that testing hypotheses across multiple policy domains is a distant goal. Collecting data on cities' policymaking (which is necessary to test whether the realization of the threat of local action spurs states to act preemptively) is no small task. I only know of one resource, the U.S. Tobacco Control Laws Database (TCLD) maintained by the American Nonsmokers' Rights Foundation,\footnote{\url{https://no-smoke.org/materials-services/lists-maps/}} that tracks both local ordinances and state preemption laws. While collecting data on cities' attempts to pass other regulations is important (especially in the policy areas that have seen the sharpest rise in preemption activity in recent years), my aims are more modest. I hope to use the TCLD's data on local tobacco regulation and state preemption over the past 40 years to test whether my three conditions hold. In the following pages, I provide further justification of the three conditions and offer concrete means of testing them in the context of tobacco preemption.

\subsubsection*{Hypothesis 1}
The first condition concerns cities' capacities to pass regulations. As described above, cities have often lacked this power. Early formulations of Home Rule, paired with narrow interpretations of cities' sphere of policymaking authority, hindered cities' ability to legislate as they saw fit. Since mid-century, many states have adopted legislative Home Rule regimes that permit a greater degree of state policymaking while allowing state courts and legislatures to intervene and deem local ordinances in conflict with state law. Today, while legislative Home Rule predominates, cities are governed by a patchwork of state legal regimes, with different implications for the likelihood of state preemption. In states where Dillon's Rule is still frequently applied by judges, cities possess limited policymaking authority. They can only legislate in policy domains that have been explicitly permitted by the state. While states can theoretically grant wide authority to cities and then preempt ordinances by curtailing city's delegated power, states that employ Dillon's Rule usually delegate narrow powers to cities in the first place. As a result, cities have little capacity to legislate and if they do act outside the scope of their delegated authority, it usually falls to the state courts to determine if the ordinance is legal \parencite{dillerIntrastatePreemption2007}.

In other states, cities' authority is structured according to the earlier formulation of Home Rule. Cities enjoy authority over matters of local concern, but the bounds of ``local concern'' is up to state judges' interpretation. Consequently, if a city ordinance is preempted, it is usually the result of a court deciding that the city acted outside of its local area of concern \parencite{dillerIntrastatePreemption2007}. In contrast, legislative Home Rule empowers cities to pass any ordinance that does not conflict with state law. While cities enjoy broad authority in such a regime, state legislatures, as discussed above, also possess the authority to intervene and curtail the bounds of local autonomy. Therefore, it is in states that govern cities with legislative Home Rule that we should expect to see the greatest level of preemption activity.

Unfortunately, given the complexities of state-local relations, there is no clear method for classifying states' legal regimes. Even the distinction between Home Rule and Dillon's Rule has been challenged \parencite{richardsonDillonRuleMars2011}. In a review of different classification schemes, \textcite{dillerIntrastatePreemption2007} notes that some scholars count 48 states with Home Rule while others count 45. Differentiating between the type of Home Rule is even trickier. One study counts 26 states with legislative Home Rule while others argue that the vast majority of Home Rule states employ legislative Home Rule. Moreover, classification disagreements do not only occur at the margin. In their analysis of the causes of state preemption, \textcite{fowlerStatePreemptionLocal2019}, citing \textcite{kraneHomeRuleAmerica2000}, only treat 10 states as using Home Rule.

Classifying the legal regimes governing cities in the states is a challenging but necessary step for measuring how the threat of local action affects states' decisions to preempt local ordinances. While states' legal regimes provide one perspective on the threat of local action, another more concrete measure is whether a city in a state has already passed a tobacco regulation. Unlike the state's legal rules governing cities, which measures the potential for action, whether a city has already regulated the sale or consumption of tobacco captures the realization of the threat. Controlling for cities' policymaking is notably absent from recent quantitative analyses of preemption \parencite{fowlerStatePreemptionLocal2019,goodmanStateLegislativeIdeology2019,flavinExplainingStatePreemption2019}, likely due to the difficulty of collecting data on cities' policies. Luckily, the TCLD systematically tracks local tobacco regulations alongside states' preemption laws. It includes the tobacco regulations of 98\% of the cities in the U.S. with more than 75,000 inhabitants, allowing me to test my hypotheses across all fifty states.

\subsubsection*{Hypothesis 2}
The second condition contends that diverging policy preferences between the state government and at least one city in the state increases the likelihood of state preemption. Struggles between cities and states over the bounds of local autonomy usually pit conservative state governments against liberal cities \parencite{rapoportBlueCitiesBattle2016,riverstone-newellRiseStatePreemption2017}, although this is not always the case.\footnote{For instance, in 2013, Oregon preempted cities from regulating genetically-modified crops and, in 2017, New York overturned New York City's tax on plastic bags \parencite{mckinleyCuomoBlocksNew2017,scharffHyperPreemptionReordering2018}.} Given the increasing number of state preemption laws targeting local economic regulations and increasing polarization between Republicans and Democrats on economic and labor issues over the past two decades \parencite{grumbachBackwatersMajorPolicymakers2018}, Republican control of government and state conservatism, unsurprisingly, play a key explanatory role in recent studies of preemption \parencite{hicksHomeRuleBe2018,riverstone-newellRiseStatePreemption2017,flavinExplainingStatePreemption2019,fowlerStatePreemptionLocal2019}.

While Republican control of government and conservatism certainly correlate with recent preemption activity, we do not yet have a clear idea of whether partisanship and ideology matter in absolute or relative terms. In other words, we do not know whether Republican control of government and conservatism inherently lead to preemption or whether the connection hinges on the relative difference between the preferences and ideology of the state and its cities. Should we expect preemption to be just as likely in North Dakota and Texas, since they are both conservative states controlled by Republicans? Or does the fact that Texas contains much more progressive municipalities mean that it is more likely to preempt local ordinances? Based on the above discussion of preemption's history, where preemption often seems to occur as a response to local actions, the latter framing seems to be a more precise formulation of the role of partisanship and ideology. It is also more generalizable, as it does not assume that certain parties and ideologies always pursue preemption.

While treating ideology and policy preferences in relative terms is important, it is also challenging. Measuring the liberal ideology of cities is difficult, especially over multiple decades. One, albeit imperfect, approach is to use the share of votes earned by Democratic presidential candidates in urban counties. Figure \ref{fig:ideology} uses this approach to display the growing ideological divide between states and cities. The dashed lines in the figure show the increasing relative support Democratic presidential candidates enjoy in the three largest counties in states that support the Democratic and Republican candidate, respectively. Over time, the largest counties have become increasingly more liberal on average than their states. This is especially true for counties in states that support the Republican presidential candidate (red dashed line).

\begin{figure}[!ht]
\caption{Average relative support of Democratic presidential candidate by geography and state partisanship, 1980-2016}
\label{fig:ideology}
\centering
\includegraphics[width=.75\textwidth]{plots/county_support}
\newline\scriptsize{This plot shows the average support for the Democratic presidential candidate in and outside the three largest counties in a state relative to the candidate's support in the rest of the state. For each state and year, the democratic candidate's vote share in and outside each state's three largest counties is divided by his or her vote share in the entire state. Annual average relative support is then calculated, grouping by whether a state supported the Democratic candidate and whether the counties are among the three largest in size. Given time constraints, the total number of votes in a county is used as a proxy for the total population of the county. County-level election returns collected from \textcite{cqpressCQVotingElections2019}.}
\end{figure}

The growing ideological divide between urban and rural areas has been well-documented elsewhere \parencite[e.g.][]{chenUnintentionalGerrymanderingPolitical2013} and cited as a cause of increasing levels of state preemption \parencite{dupuisCityRightsEra2018,hicksHomeRuleBe2018}. As Democratic voters cluster in urban areas, they increase the liberalism of cities relative to states and reduce the number of state-level Republicans representing urban areas. As a result, cities may be more willing to enact more liberal policies that increasingly diverge from the interests of state elected officials, increasing the likelihood that the state will preempt progressive local regulations. In the context of state preemption of tobacco regulations, Figure \ref{fig:ideology} suggests that ideological divergence between rural and urban counties began in the late 1980s, around the time states began to preempt local tobacco ordinances. Interestingly, until about 2000, the ideological gulf between cities and states in states that supported the Democratic presidential candidate was consistently larger than the within-state ideological divide in Republican states. Therefore, if local tobacco regulations occurred in more progressive cities (as measured by support for the Democratic presidential candidate), we might expect to see preemption occurring in both Democratic and Republican states.

Figure \ref{fig:party}, which charts the number of state governments under unified Republican control from 1980 to 2017, also implies that the preemption of tobacco laws may not have been solely driven by the actions of conservative, Republican-controlled governments. The figure shows that just as the 2010s have been a decade of Republican dominance in the states, the 1980s and early 1990s were a period of Democratic control of state governorships and legislatures. Since tobacco preemption began in earnest before Republicans gained widespread power in the states in the 1994 elections, preemption seems to have occurred in many Democratic states.

The above discussion suggests that we need to move beyond Republican control of government and conservative ideology as intrinsically linked to state preemption of local laws. While recent decades' increasing polarization and diverging preferences between rural and urban areas means that conservative, Republican-controlled states are often the source of recent preemption activity, it does not seem that this has always been the case. The ideological and partisan geography of the U.S. in the late 1980s and 1990s suggests that preemption must have been occurring in Democratic states. It will take getting the TCLD data in-hand for me to determine whether this circumstantial evidence holds up to further scrutiny.

\begin{figure}[!ht]
\caption{Number of state governments under unified partisan control, 1980-2017}
\label{fig:party}
\centering
\includegraphics[width=.75\textwidth]{plots/party_control}
\newline\scriptsize{Note: Excludes Nebraska. Sources: \textcite{jordanCorrelatesStatePolicy2017,nationalconferenceofstatelegislaturesStatePartisanComposition2019}.}
\end{figure}

\subsubsection*{Hypothesis 3}
My third and final hypothesis states that preemption is more likely when relevant interest groups are mobilized at the state level.

\singlespacing
\pagebreak
\printbibliography
